DRAFT \\ Hamiltonian mechanics is conservation of information

### 2017-02-16 2017HamiltonianInformation/hamiltonianInformation.tex
Updating colors for figures

\newif\ifjournal
\journaltrue
%\journalfalse

@@

\title{From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics}
\author{Gabriele Carcassi, Christine A. Aidala, David J. Baker, Lydia Bieri}

\ifjournal
	\institute{G. Carcassi, C. A. Aidala \at
		Physics Department, University of Michigan, Ann Arbor, MI 48109 \\
		\email{carcassi@umich.edu}
		\and
		D. J. Baker \at
		Philosophy Department, University of Michigan, Ann Arbor, MI 48109
		\and
		L. Bieri \at
		Mathematics Department, University of Michigan, Ann Arbor, MI 48109 \\
	}
\else
	\affiliation{University of Michigan, Ann Arbor, MI 48109}
\fi

\ifjournal
	% As the page margins are so large, the full title does not fit
	\titlerunning{From physical principles to classical and quantum particle mechanics}
\fi

@@

\begin{abstract}
The aim of this work is to show that particle mechanics, both classical and quantum, Hamiltonian and Lagrangian, can be derived from few simple physical assumptions. Assuming deterministic and reversible time evolution will give us a dynamical system whose set of states form a topological space and whose law of evolution is a self-homeomorphism. Assuming the system is infinitesimally reducible, that giving the state and the dynamics of the whole system is equivalent to giving the state and the dynamics of its infinitesimal parts, will give us a classical Hamiltonian system. Assuming the system is irreducible, that giving the state and the dynamics of the whole system tells us nothing about the state and the dynamics of its substructure, will give us a quantum Hamiltonian system. Assuming kinematic equivalence, that studying trajectories is equivalent to studying state evolution, will give us Lagrangian mechanics and limit the form of the Hamiltonian/Lagrangian to the one with scalar and vector potential forces.
\keywords{Hamiltonian mechanics \and Lagrangian mechanics \and Quantum mechanics \and Fundamental assumptions}
\end{abstract}

@@

The past century has seen a resounding success in the use of mathematics in fundamental physics. In many cases, mathematical ideas were what allowed progress in theoretical physics and ultimately led to experimental discoveries. While these ideas are indeed useful---anything that can advance our understanding is welcome---they cannot, however, be the whole story.

The issue is that mathematical structures are not enough to characterize what physical system is being studied. For example, the same mathematical framework for linear circuits can be used to describe electric, hydraulic, thermal or mechanical systems~\cite{Roland,Barron,Borutzky}. The mathematical treatment is the same because it captures aspects of the description that are common to all cases; that indeed is its power. But for the very same reason, we cannot infer what system is being described if all we are given is the mathematical model. In the jargon of linguistics/computer science/philosophy~\cite{Freidin,Friedman}: math captures the syntax (the relationships between the objects) but does not capture the semantics (the meaning of each object).

The real trouble comes when the mathematical description essentially becomes the foundation of a physical theory. While some branches of physics, such as Newtonian mechanics, thermodynamics and special relativity, are founded on physical laws or assumptions, others, such as Hamiltonian, Lagrangian and quantum mechanics, start by setting their mathematical structure. Nothing tells us why there should be  a Lagrangian function whose path integral is minimized during the motion, or when a state is described by conjugate pairs $(q^i, p_i)$ instead of a vector in a complex inner product space. Those are taken as given.

This means that if we have a system in front of us, we do not have a rigorous way to conceptually establish whether the system is Hamiltonian or not, whether it is classical or quantum. We have some heuristics we can apply, but, at the end the day, a system is Hamiltonian because it follows the Hamiltonian framework.\footnote{The lack of a properly defined, and uniquely accepted, semantics is of course very evident in quantum mechanics with its numerous interpretations.}

Unsatisfied with this situation, we started creating a sort of math-to-physics dictionary so that each mathematical concept could be clearly associated to a crisp physical concept. As we pushed further back toward the basic mathematical definitions, we ended up with physical concepts of a more general nature and found that, in some cases, a single physical idea could have more than one mathematical consequence. That is: by putting physics at the center, we could achieve a unification of ideas that was not evident in the math alone. The unexpected success of the approach led us to think that, perhaps, the picture we were developing was a better framework to understand and link different areas of basic science and mathematics. This work is the systematized result of this endeavor.

The main goal is to show that it is possible to derive classical/quantum Hamiltonian/Lagrangian particle mechanics starting from few physical assumptions. We do \emph{not} claim that we have perfectly achieved this result, with no room for improvement. We \emph{do} claim that the whole architectural framework is solid enough to be convinced that not only is this indeed possible, but it allows a more unified picture that facilitates intuitive connections within and across different areas of physics and mathematics. It enables a consistent narrative across several domains of knowledge.

Rederiving classical, quantum and (to some extent) relativistic mechanics within the same work is therefore crucial to show that the concepts we develop are truly general, that they can be applied to many different cases. If we could derive classical Hamiltonian mechanics on assumptions that are inapplicable to the quantum case, what good would it do? Quantum systems are Hamiltonian systems, so clearly we wouldn't have understood what makes a system Hamiltonian. This breadth of applicability is what makes the work valuable.

As a brief overview, there are four assumptions we will consider. Assuming deterministic and reversible evolution gives us a dynamical system: states and a law of evolution. Additionally assuming infinitesimal reducibility, studying the whole system is equivalent to studying its infinitesimal parts, gives us Hamiltonian mechanics. Further assuming kinematic equivalence, trajectories in space tell us everything about the state of the system, gives us Lagrangian mechanics and constrains the motion to massive particles under scalar/vector potential forces. Relativistic motion arises when studying time dependent laws of evolution with no further assumptions. Alternatively, as the second assumption we can instead take irreducibility, studying the whole system tells us nothing about its parts, which leads to quantum (Hamiltonian) mechanics.

As a more detailed overview: in section \ref{sec:fundamental_model} we describe in general what it means to study a physical system and how the assumption of determinism and reversibility plays into it. We develop an abstract conceptual model that we use when discussing the later assumptions.

In section \ref{sec:dynamical_systems} we show how the physical concept of physical distinguishability leads to the mathematical concept of topological spaces. Deterministic and reversible evolution is an invertible continuous map (i.e. a homeomorphism) as it needs to preserve what is physically distinguishable.

In section \ref{sec:reducibility} we introduce the notion of a classical material, one for which we assume infinitesimal reducibility. The state of an arbitrary amount of material is a distribution over the states of the infinitesimal parts, the classical particles. As such a distribution must be invariant under the arbitrary choice of units, we see that for each variable $q^i$ that defines a unit there is a corresponding variable $k^i$ defined on the inverse of that unit. Classical particle states can be identified by conjugate pairs of state variables $(q^i, p_i)$ (i.e. a point in the cotangent bundle $T^*\mathcal{Q}$) and the number of possible initial conditions for each pair is given by $dq^i \wedge dp_i$ (i.e. the canonical symplectic form $\omega$). During evolution, the distribution (and marginal distributions) must be mapped point by point which leads to Hamiltonian mechanics (i.e. a symplectomorphism on $T^*\mathcal{Q}$).

In section \ref{sec:relativity} we show that a relativistic version of Hamilton's equations arises naturally when we consider time dependent laws of evolution for distributions of classical material. This also gives the notion of deterministic and reversible evolution a characterization that is independent of any transformation that mixes time and state variables $(q^i, p_i)$. Moreover it gives us a classical equivalent for anti-particle states.

In section \ref{sec:Lagrangian} we introduce the assumption of kinematic equivalence, studying the motion in space is equivalent to studying the evolution of the state. Under that assumption there is a link between the state variables $(q^i, p_i)$ and the kinematic variables $(x^i, u^i)$ (i.e. position and velocity). This link allows us to express the number of possible initial conditions for a degree of freedom (d.o.f.) in terms of $dx^i$ and $du^i$, leading to the metric tensor $g$. Such an expression also allows us to constrain the Hamiltonian to the one for massive particles under scalar/vector potential forces and an equivalent Lagrangian formulation.

In section \ref{sec:quantum_prelude} we introduce some ideas in classical mechanics to more gently transition to quantum mechanics. We see how distributions over phase space already have features that are typically associated with quantum systems and we will show how to construct a classical uncertainty principle that helps to build intuition used in the later section.

In section \ref{sec:irreducibility} we introduce the notion of an irreducible material, one for which studying the whole distribution does not tell us anything about the motion of its infinitesimal parts, the fragments. A unit amount of this material will correspond to a quantum particle. As this means that the states of such a material are invariant under permutations of the configurations of the fragments, we show that the state space of an irreducible material is a complex vector space. These states can still be compared to each other in a way that gives the state space an inner product. Such comparisons will be preserved by deterministic and reversible evolution which leads to Schroedinger's equation (i.e. a unitary transformation).

In section \ref{sec:discussion} we present a few higher level remarks about the work.

\section{Organization and style}

The overall layout of this work is a balance between the need to show that the derivation follows, which requires starting from basic principles and proceeding using formal arguments, and the need to give an intuitive understanding, which often requires starting from specific cases and then generalizing.

The deductive aspect is formally organized into the following:
\begin{description}
  \item[Assumptions] these characterize the physical system we are studying and constitute the premise of our discussion. A \textbf{rationale} follows each assumption, which uses physical and sometimes philosophical arguments to motivate why (or not) such an assumption makes sense in a particular case.

  \item[Propositions] these capture the properties of our physical objects into mathematical language. A \textbf{justification} follows each proposition to show the necessity of such a characterization. While it is a formal argument, it is not a strictly mathematical proof as it contains physical arguments.

  \item[Theorems] these are pure mathematical statements that are used to simplify propositions. A mathematical \textbf{proof} follows each theorem. No mathematical breakthrough should be expected as we mostly use well known results from different fields. Theorems are grouped in the appendix.
\end{description}

As these sections are clearly demarcated, they allow us to make sure that each justification only references past propositions and avoid circular arguments. It also makes it easier to read the work at the desired level of detail. Interspersed between each step of the more formal treatment, there are numerous discussions that conceptually explain motivations and results in a more approachable way. These can use more concrete and intuitive examples which may, in some cases, make forward references. As both the formal and more intuitive parts need to cover the same material, this will inevitably lead to some repetition.

The work uses ideas and tools from a broad range of mathematical and scientific subjects: topology~\cite{LeeTM}, measure theory~\cite{Halmos}, differential geometry (symplectic and Riemannian)~\cite{Lee}, group theory and vector spaces~\cite{Lang,Young}, statistics~\cite{Grimmet}, information theory~\cite{Pierce}, Hamiltonian mechanics, Lagrangian mechanics~\cite{classical_dynamics}, quantum mechanics~\cite{Landau,Weinberg}, thermodynamics~\cite{Van Ness}, special relativity~\cite{French} and so on.\footnote{The references given provide the reader a list of resources consistent with the development of this work.} Using ideas from all these areas is necessary and valuable, but it does create a number of problems.

The first one is terminology and prerequisite knowledge. Since we touch numerous areas, we cannot expect all readers to be equally comfortable with all the technical terms. Yet we cannot introduce all of them either. The compromise we adopted is to keep the informal discussion more accessible while keeping the formal derivation more precise, without needing more than the basic definitions and few major results within each specific area. This may leave an expert in a particular field not fully satisfied but, as stated in the introduction, our goal is to lay out the general picture.

The second problem is notation. As each field has developed its own set of conventions, the common usage is now an empty set. We ended up with a compromise that hopefully feels natural enough to a broad audience of physicists, mathematical physicists and philosophers of physics. Within formal parts we may use notation and terminology more specific to the appropriate area of math.

We have tested the prose on a number of colleagues and students in physics, philosophy and mathematics to confirm its accessibility at different levels and believe we have struck a reasonable compromise.

\section{On studying a physical system}
\label{sec:fundamental_model}

% Status: Conceptual work done. Text ready. Incorporated Dave, Isaac and Christine feedback

Our first task is to develop a conceptual model that applies to all realms of physics we'll be considering: classical, statistical and, later, quantum mechanics. We will take the standard picture of system plus environment and extend it to differentiate between the state of the system (i.e. the aspects under study) from the unstated part of the system (i.e. the aspects missing from our description).

We will assume that the state evolves according to a deterministic and reversible law (i.e. for each present state there is one and only one future state). While the unstated part does not influence the state evolution, we'll see how it constrains what states are available and what type of description can or cannot be given to the system.

\subsection{States and their evolution}

We start by fixing a \emph{physical system}, meaning something we can interact with and perform measurements on (e.g. a planet, a fluid, ...). We call \emph{environment} everything else. We set what particular aspect we want to study (e.g. the motion around a star, the flow in a pipe, ...). We call \emph{state} a physically distinguishable configuration of the aspect under study at a particular time (e.g. position/momentum of center of mass, velocity field, ...). Since the state does not, in general, exhaust the description of the system, a part remains \emph{unstated}, and as such we'll call it, for lack of a better word (e.g. the chemical composition, the motion of each of its molecules, ...). Note that the environment plays an essential role here as it's what allows us to define two states as physically distinguishable: we can find an external process (i.e. part of the environment) whose outcome changes depending on the different state. As such processes are what we can use to perform measurements, we consider the experimental apparatus (and us performing measurements) independent of the system, part of the environment.\footnote{This is true even in the case of general relativity, where we can imagine multiple researchers on small spaceships collecting data without greatly influencing the motion of stars and planets. The case where the physical system is the whole universe and there is no environment presents practical problems  and conceptual challenges that the current physical theories do not seem to be equipped to address, and therefore will be absent from our discussion. For example, what physical device can we use to store and process the state of the whole universe to make predictions and compare? How do we define physically distinguishable? Do the physical laws determine which measurements we are going to make and does that limit what is actually distinguishable?}

In this context, we call the evolution of the system \emph{deterministic} if the state at a given time uniquely identifies states at future times, and \emph{reversible} if it uniquely identifies states at past times. While this is a common enough definition, we need to be clear how this applies to the unstated part. Note that the concept of determinism outlined here is context dependent because the state itself represents only the part of the system that we choose to (or can) describe. In this sense, the unstated part is \emph{always} non-deterministic (and non-reversible) in the sense that the state of the system does not determine its evolution. For example, suppose we define the state as position and momentum of the center of mass of a cannonball. Suppose that the evolution is deterministic on that state. What does it tell us about the cannonball temperature, or about the motion of each of its atoms? Nothing. In this sense, the unstated part is non-deterministic and non-reversible. Could we extend the state and the laws of evolution to account for temperature? Yes, but that would be a different evolution defined on a different state. Does it mean the unstated part is always evolving randomly? Not at all. The temperature may remain constant throughout the motion of the cannonball. Yet we wouldn't know, since we are not studying it: the evolution is deterministic and/or reversible only as far as the state is concerned. With this in mind, we will restrict ourselves to the cases where the following is valid:

\begin{assump}[Determinism and reversibility]\label{ass:determinism}
The state of the physical system under study undergoes deterministic and reversible evolution.
\end{assump}

\begin{rationale}
As it is an assumption, we first need to discuss when it is valid. More specifically, we need to understand that the non-deterministic/non-reversible evolution of the unstated part plays as much of a fundamental role as the deterministic/reversible evolution of the state. In fact, the non-deterministic part contributes in determining what states are available to the system.

Suppose we study the motion of a cannonball; its state under gravitational and (inertial) inertial forces will be properly described by the position and momentum of the center of mass. While light and air molecules may scatter off its surface unpredictably, its trajectory is not greatly affected as it is a massive rigid body. Suppose we study the motion of a small particle, small enough that the random scattering does influence the trajectory and it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the states have changed from ``pure" to statistical ensembles. In other words, the set of states must be closed under both the deterministic evolution of the state and the non-deterministic evolution of the unstated part. If the Brownian motion is not negligible, we do not end in a well defined position/momentum pair, even if we start from one.

A similar more drastic effect: consider a book and its motion under gravitation and inertial forces, its state being the position and momentum of the center of mass. As we increase the temperature of the air around the book, its motion remains unaffected until, at some point, the book burns. Clearly, the non-deterministic evolution has pushed one of the states outside the set of states, to the point that the system is no longer recognizable.

As we have hinted, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is, given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the evolution of the parts are not as they fall within the unstated part. Note that we \emph{cannot} assume trajectories and states are always defined for the unstated part, as this includes also the unknown unknowns. We will return to this aspect when discussing quantum systems.

It should also be clear that what constitutes state and unstated part does not depend only on the system under study, but also on the processes we are considering. In some circumstances, the chemical composition of a fluid may be relevant, in others it may not. The choices of environment, state and unstated part are not independent from each other. By choosing a particular set of states, we are not only saying that the state evolution is well approximated by a deterministic/reversible map from initial to final state, but we are also saying that the non-deterministic/non-reversible evolution of the unstated part does not change the nature of the system, and processes that do not satisfy these conditions are not under consideration.

As with all assumptions, we should also ask whether it is necessary. That is, could we define a set of physically distinguishable states and yet have no deterministic and reversible processes defined on them? The claim is that this assumption is indeed needed, as without it we cannot properly define states or write useful physics laws. We can provide different arguments that point in the same direction.

First, to be able to identify the system, we must be able to tell it apart from anything else. Intuitively, we can distinguish between two chairs because we can move the first to another room and sit on it without having touched the second. We can manipulate the state of the first system without affecting the second, and vice-versa. So, to identify a system it has to be sufficiently isolated from everything else. This means that the system future and past states are with good approximation determined only by its own state: the state undergoes deterministic and reversible evolution.

Second, the aim of physics is to write laws that can be used to make predictions that can be validated experimentally. If I drop an anvil from a tower, it will accelerate at $9.81$~m/s$^2$; if I want the anvil to reach the ground at $x$~m/s I have to drop it from $y$~m. To the extent that we want to make predictions in time, we need to have a correspondence between initial and final states.

Third, operationally we must reliably prepare and measure states. That is, we need a process for which the input settings of our preparing device determine the outgoing state of the system; and a process for which the incoming state of the system can be reconstructed by the output of the measuring device. That is, our system must, at least in some cases, be able to participate in a deterministic and reversible process with the preparing and measuring device. Without it we wouldn't be able to calibrate our experimental apparatus.

Fourth, to be able to ascribe a property to a system we need to claim that, at least for a finite interval of time, the system either held or did not hold such property. That is, there is a deterministic and reversible process for that finite period of time for such property.

This link between state definition and deterministic processes should not be too surprising as the state, in the context of control theory and dynamical systems theory, is often defined as \emph{the set of variables needed to determine the future evolution of the system}~\cite{Katok,Sontang}. As we saw before, this applies also to statistical processes: the distribution (the ensemble) as a whole can indeed be calculated, measured and prepared. We can also describe the evolution of each element of the distribution provided that: we have a way to isolate it and study it under deterministic and reversible motion (so that we can define microstates); the non-deterministic motion does not alter the system (the set of microstates is preserved by the evolution).

As with many assumptions, we should stress that it's an idealization: it can never be completely achieved in practice. A system can be prepared or measured up to a certain level of precision. Perfect determinism and isolation of a system is impossible both practically (e.g. black-body radiation, gravity, ...) and conceptually (e.g. if the system is perfectly isolated, we cannot interact with it: how can it be physically distinguished?). It's a simplifying assumption that can only be taken if the environment and the internal dynamics of the system interact in such a way that they little affect and are little affected by the aspect we are studying. As we saw before, for example, assuming that the state consists of the position and momentum of the center of mass requires assuming that the Brownian motion of the body is negligible.

Yet, this is a fundamental assumption in the sense that it is needed. If a particular set of states does not satisfy deterministic and reversible evolution under certain conditions, what we do is to keep at it until we find a set that does. That is, we work to restore the assumption. Finding new sets of states with new laws of evolution is, in fact, what leads to new physics. We therefore call \emph{fundamental model of physics} the triad of state, unstated part and environment, with the assumption that the state may undergo deterministic and reversible evolution, and the unstated part undergoes non-deterministic non-reversible evolution that does not alter the set of states.
\end{rationale}

\section{States and state space}
\label{sec:dynamical_systems}

% Status: First subsection. Conceptual work done. Text ready. Incorporated Christine feedback
% Second subsection. Conceptual work almost done. Need to finish the text

%Open questions: smoothness (obviously required to define densities later, is it required in general? What does it means that physical processes/measurements don't give a smooth topology?)

We now proceed to characterize states and physical distinguishability in more precise terms so that we can capture their description mathematically. What we'll see is that the outcomes of all physical processes that can be used to gain information about the system, that can be used to perform a measurement, induce a topology on the set of states that is at least Hausdorff. That is, physical distinguishability is mathematically captured by topological distinguishability. Deterministic and reversible evolution will then preserve the topology, and they will be mathematically captured by self-homeomorphisms.

We'll focus on state spaces that can be described by a set of independent state variables, either discrete or continuous, and see what can be said in general on the evolution of state variables.

\subsection{States and topology}

As the term ``measurement" has become particularly loaded, let's first characterize what we mean by physical distinguishability in our context.

Consider the motion of a cannonball under inertial and gravitational forces. Light will scatter off of it; as it lands, the ground will be deformed and the impact will make the temperature rise slightly. Those external physical processes, which happen no matter what we do, can be used to distinguish the motion of the cannonball as their outcomes are correlated. Therefore we can learn the position by looking at the reflected light, and learn the final kinetic energy by looking at the deformation of the ground. In principle, any external process that has a correlation with the states under study can be used to perform a measurement, and any measurement is based on such a process. That is, for us a measurement is simply a physical process that we can use to distinguish states. Setting up an experiment means choosing a particular process with desired outcomes and forcing the system under study to interact with it one or more times. After that, there is no special role played by the ``observer" in making outcomes come about.

This external process may be quite complicated: when a particle enters a calorimeter, a shower of particles is produced, photons are captured and are directed to photomultipliers, a current is read out, the current is then digitized, and so on. Some processes interfere with the system, they affect its dynamics, and others are destructive, the system no longer can be described by the original set of states. A tracking chamber is an example of the first (the magnetic field curves the motion of a charged particle); burning a substance to determine its caloric content is an example of the latter. Therefore intimate knowledge of the process is always needed to ensure that one makes the proper link between outcomes and the \emph{original} states, and properly accounts for systematic uncertainties that would skew that link.

Repeatability is also fundamental. First, to make sure the process is indeed correlated to the states. Second, because ``a single take does not a measurement make." One has to gather enough statistics. Note that the number of takes influences the outcomes: with greater statistics the precision and number of distinguishable cases increases. Therefore the processes, as we defined them, may require repeated interactions with similarly prepared states. They may even be a combination of different kinds of interactions that, taken all together, create a set of distinguishable outcomes. But in the end, however complicated it is, the conceptual model remains the same: each process has a set of possible outcomes, and each possible outcome will be associated with a set of states consistent with that outcome. For example, if the cannonball deformed the ground by this amount then its kinetic energy at impact was within this range; if the electron follows a certain path, then its state is among the ones that have spin $+1/2$.

Note that the outcome of the process (e.g. the electron followed a particular path) is conceptually distinct from the set of states compatible with it (the states associated with the path) and the way we label them (spin=$+1/2$). Yet,  we'll conflate the concepts as this makes the language closer to experimental physics (``the outcome of the measurement is spin=$+1/2$"). Context will allow us to recover the distinction.

However precise our measurements are, we can only gather a finite amount of statistics and each outcome is expressed by a finite set of digits; therefore, the set of outcomes is countable.\footnote{The information provided by the process as measured by Shannon's entropy~\cite{Shannon,Jaynes} is finite.} Note that different outcomes for the same process can overlap. For example, $4.12 \pm 0.05$ cm and $4.13 \pm 0.05$ cm are both legitimate possible outputs of the same measurement device. But since all states must be distinguishable, given two arbitrary states there must be a process precise enough to tell them apart. That is, the potential outcomes associated with the two states do not overlap. For example, $4.12 \pm 0.0005$ cm and $4.13 \pm 0.0005$ cm do not overlap anymore.

We can also conceptually combine two different processes into a single one.\footnote{That is, take the logical \textsc{and} between outcomes of different processes.} That is, having a way to measure quantity $x$ and a way to measure quantity $y$ gives us a way to measure the combination $(x,y)$. For ensembles, if we can measure the marginal distribution $\rho_x(x)$ and the marginal distribution $\rho_y(y)$, we know the joint distribution $\rho(x,y)$ has to be compatible with both. Note, though, that this does not provide a way to fully measure $\rho(x,y)$ as we know nothing about the correlation between the two variables.\footnote{The quantum case is similar: we measure marginal distributions and rule out states that are incompatible with them.} Formally, the states compatible with the outcomes of the combined process will be the intersections of the states compatible with each pair of outcomes of the original processes.

We can also coarsen a single process.\footnote{That is, take the logical \textsc{or} between outcomes of the same process.} That is, having a way to measure $(x,y)$ gives us a way to measure $x$ alone (or any $f(x,y)$). Formally, the outcomes of the coarsened process are given by performing the union of some outcomes of the original process.

This model maps very naturally to a topological space. The states are elements of a set and the physical outcomes provide a topology on that set. The fact that two elements of the set can be distinguished requires the space to be Hausdorff.

\begin{prop}\label{prop:state_topology}
The state space $\mathcal{S}$ of a physical system is a Hausdorff topological space.
\end{prop}

\begin{justification}
We claim $\mathcal{S}$ is a set. Each state is well defined as it is physically distinguishable. The collection of all possible states forms a set.

We claim $\mathcal{S}$ has a topology $\mathsf{T}$. Consider the set of all possible physical outcomes associated with all physical processes. Each possible outcome is associated with a set of states that are compatible with that outcome. Let $\mathsf{T}$ be the set of all sets associated with all physical outcomes. $\mathcal{S} \in \mathsf{T}$ and is associated with the outcome ``the system exists." $\varnothing \in \mathsf{T}$ and is associated with the outcome ``the system doesn't exist." Let $V_1, V_2 \in \mathsf{T}$. Then, by definition, there exists a process $P_1$ that admits $V_1$ as an outcome and a process $P_2$ that admits $V_2$ as an outcome. Consider the process $P$ that combines the outcomes of $P_1$ and $P_2$ with a logical \textsc{and}. This always exists physically as we can prepare the same state multiple times and let it interact with each process separately. $P$ will have a possible outcome $V$ corresponding to the case where $P_1$ gave outcome $V_1$ and $P_2$ gave outcome $V_2$. The states compatible with $V$ must be in both $V_1$ and $V_2$, that is $V = V_1 \cap V_2$. Therefore $\mathsf{T}$ is closed under intersection. Let $V_1, V_2 \in \mathsf{T}$. If they are physically distinguishable, then there exists a physical process $P$ that admits both as outcomes. Given $P$, we can always construct the process $P_0$ that combines $V_1, V_2$ with a logical \textsc{or} into a single outcome $V$, by ``forgetting" which of the two was given. The states compatible with $V$ must be in either $V_1$ or $V_2$, that is $V = V_1 \cup V_2$. Therefore $\mathsf{T}$ is closed under union. $\mathsf{T}$ is a topology by definition.

We claim that $\mathsf{T}$ is Hausdorff. Let $\mathcal{s_1}, \mathcal{s_2} \in \mathcal{S}$. As states are physically distinguishable, there must exist a physical process with two possible outcomes $V_1, V_2 \in \mathsf{T}$ for which $s_1 \in V_1, s_2 \in V_2, V_1 \cap V_2 = \varnothing$. $\mathcal{S}$ is Hausdorff by definition.
\end{justification}

Note that the arguments that led to the topological space had nothing to do with states per se, just that they are physically distinguishable. States, though, are not the only objects with that property. In fact, any element of a set of physical objects (e.g. forces, physical properties such as mass or charge, time) needs to be physically distinguishable to be well defined. We can generalize the above justification: any set of physically distinguishable elements is a topological space.

\begin{prop}\label{prop:topology}
	Any set $\mathcal{S}$ of physically distinguishable elements is a Hausdorff topological space.
\end{prop}

\begin{justification}
	Same justification as in \ref{prop:state_topology} with ``state" replaced by ``element" of the set $\mathcal{S}$.
\end{justification}

With our state space defined, deterministic and reversible evolution corresponds to a bijective map between initial and final states. But that's not enough. A physical process that can distinguish final states can be used, together with the evolution, to distinguish initial states. We prepare the initial state, let it evolve deterministically, measure the final state and use reversibility to convert the measure to one on the initial state. Therefore, not only the evolution is a bijection, but it also maps outcomes to outcomes. That is: the topology is mapped and preserved by the deterministic and reversible evolution because physical distinguishability must remain unchanged. The evolution is then a self-homeomorphism on the state space.

\begin{prop}\label{prop:homeomorphism}
A deterministic and reversible evolution map is a self-\journal{\break}homeomorphism on the state space $\mathcal{T}_{\Delta t}:\mathcal{S} \rightarrow \mathcal{S}$.
\end{prop}

\begin{justification}
We claim that $\mathcal{T}_{\Delta t}$ exists. The system is deterministic: given an initial state $\mathcal{s} \in \mathcal{S}$ there exists a well defined final state $\mathcal{T}_{\Delta t}(\mathcal{s})=\hat{\mathcal{s}} \in \mathcal{S}$.

We claim $\mathcal{T}_{\Delta t}$ is continuous. Let $U \subseteq \mathcal{S}$ represent an outcome of a process $P$ on the final states. $U$ is an open set in the (final) state space topology by definition. Consider the process $P_0$ that first evolves the initial states and then distinguishes the final state with $P$. $P_0$ is a process that distinguishes initial states. The set of initial states compatible with $U$ are $\mathcal{T}_{\Delta t}^{-1}(U)$. $\mathcal{T}_{\Delta t}^{-1}(U)$ is an open set in the (initial) state space topology by definition. $\mathcal{T}_{\Delta t}$ is a continuous map.

We claim that $\mathcal{T}_{\Delta t}$ is a bijection. The system is reversible: there exists a map $\mathcal{T}_{-\Delta t}:\mathcal{S} \rightarrow \mathcal{S}$ that returns the initial state given the final state. $\mathcal{T}_{-\Delta t} \circ \mathcal{T}_{\Delta t} = \mathcal{T}_{\Delta t} \circ \mathcal{T}_{-\Delta t} = \id_{\mathcal{S}}$ as mapping forward and then backward or backwards and then forward must return the original element. $\mathcal{T}_{\Delta t}$ admits $\mathcal{T}_{-\Delta t}$ as an inverse. $\mathcal{T}_{\Delta t}$ is a bijection.

We claim that $\mathcal{T}_{\Delta t}$ is a self-homeomorphism as it is a continuous bijection.
\end{justification}

Again, we note that the arguments that lead to continuity had nothing to do with states per se, just that the relationship is between physically distinguishable objects. We can then generalize the above justification.

\begin{prop}\label{prop:continuity}
	A map $f:\mathcal{S_1} \rightarrow \mathcal{S_2}$ between two sets of physically distinguishable elements $\mathcal{S_1}$ and $\mathcal{S_2}$ is a continuous map.
\end{prop}

\begin{justification}
	Same justification for continuity as in \ref{prop:homeomorphism} with ``initial states" and ``final states" replaced by ``elements" of $\mathcal{S_1}$ and $\mathcal{S_2}$ respectively.
\end{justification}

The generality of this result explains why in physics one always assumes functions to be ``well behaved." As the result was derived from our notion of physical distinguishability, this is not a matter of practical convenience. Suppose we were able to prepare a force field that was zero everywhere in space except at a single point. This gives us a way to tag a specific point. But it also allows us to create an outcome compatible with only a single point. A finite precision measurement of the force provides us infinite precision of space, which we ruled out.\footnote{This assumes that we are able to position the probe perfectly, which we could do if we were able to manipulate forces at that precision.} This is what the math is telling us, that if we claim that outcomes can't distinguish isolated points (i.e. standard topology on $\mathbb{R}^n$) then neither can maps. It is physical consistency that limits us to continuous functions.

\begin{figure}
	\includestandalone[width=\columnwidth]{pictures/continuousMap}
	\caption{A continuous function always maps finite precision knowledge in the vertical axis (i.e. an open set) to finite precision knowledge in the horizontal axis. A discontinuous function may map to infinite precision (i.e. a closed set). If infinite precision identification is ruled out (i.e. use of standard topology) then discontinuous functions must also be ruled out.}
	\label{fig:continuity}
\end{figure}

>>>>

\newif\ifjournal
%\journaltrue
\journalfalse

@@

\title{DRAFT \\ Hamiltonian mechanics is conservation of information}
\author{Gabriele Carcassi, Christine A. Aidala}

\ifjournal
	\institute{G. Carcassi \at
		University of Michigan, Ann Arbor, MI 48109 \\
		\email{carcassi@umich.edu}           %  \\
	}
\else
	\affiliation{University of Michigan, Ann Arbor, MI 48109}
\fi

%\ifjournal
%	% As the page margins are so large, the full title does not fit
%	\titlerunning{From physical principles to classical and quantum particle mechanics}
%\fi

@@

\begin{abstract}
\textbf{This manuscript is a work in progress.} Ideas are constantly reshaped to find more precise and elegant arguments. It is provided as is to stimulate discussion.  \textbf{Make sure you have the latest version from http://assumptionsofphysics.org}

In this work we show that the canonical transformations of classical Hamiltonian mechanics are exactly the transformations that preserve information.
\end{abstract}

@@

\section{Something}

\begin{prop}
	Transported coordinates are canonical. Distributions are function of the point and are transported.
\end{prop}

$(T^*\mathcal{Q}, \omega)$, $(q^i, p_j)$ set of canonical coordinates, $F: T^*\mathcal{Q} \rightarrow T^*\mathcal{Q}$ such that $\omega = F^* \omega$ where $F^*$ is the pullback. $\hat{q}^i = q^i \circ F^{-1}$ and $\hat{p}_i = p_i \circ F^{-1}$.

$F^* (\sum d\hat{q}^i \wedge d\hat{p}_i) =\sum F^* (d\hat{q}^i \wedge d\hat{p}_i) =  \sum (F^* d\hat{q}^i)\wedge (F^* d\hat{p}_i) = \sum d(\hat{q}^i \circ F)\wedge d(\hat{p}_i \circ F) = \sum d(q^i \circ F^{-1} \circ F)\wedge d(p_i \circ F^{-1} \circ F) = \sum dq^i\wedge dp_i = \omega = F^* \omega$. $\omega =  \sum d\hat{q}^i \wedge d\hat{p}_i$. $(\hat{q}^i, \hat{p}_j)$ is a set of canonical coordinates.

$\rho(q^i, p_j) : \mathbb{R}^{2n} \rightarrow \mathbb{R}$ integrable and differentiable. $\rho (q^i, p_j) \, dq^n \wedge dp_n = \rho (q^i, p_j) \, \omega^n = \rho (q^i(\hat{q}^i, \hat{p}_j), p_j(\hat{q}^i, \hat{p}_j)) \, d\hat{q}^n \wedge d\hat{p}_n$. $\rho$, defined on the symplectic structure, transforms like a function. We can write $\rho : (T^*\mathcal{Q}, \omega) \rightarrow \mathbb{R}$.

$\hat{\rho} : (T^*\mathcal{Q}, \omega) \rightarrow \mathbb{R}$ such that $\hat{\rho} = \rho \circ F^{-1 }$. $\hat{\rho}(\hat{q}^i, \hat{p}_j) = \hat{\rho} \circ (\hat{q}^i, \hat{p}_j)^{-1} = \rho \circ F^{-1} \circ (q^i \circ F^{-1}, p_i \circ F^{-1}) ^ {-1}= \rho \circ F^{-1} \circ F \circ (q^i, p_j) ^ {-1} = \rho \circ (q^i, p_j) ^ {-1} = \rho (q^i, p_j)$.

\begin{prop}
	Canonical transformations transport marginal distributions and preserve independence.
\end{prop}

\begin{prop}
	Canonical transformations transport mutual information.
\end{prop}

\begin{thebibliography}{0}

\end{thebibliography}
